---
title: "Geographic Disparities in Mental Health Care Access"
subtitle: "A county-level analysis of availability and affordability across California"
author: "Pallavi Singh"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    smooth-scroll: true
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
library(ggrepel)
library(tigris)
library(sf)
library(viridis)
library(patchwork)
library(gt)
library(plotly)

options(tigris_use_cache = TRUE)
theme_set(theme_minimal(base_size = 13))

pal <- c(Urban = "#2563EB", Micropolitan = "#F59E0B", Rural = "#059669")
```

## Overview

Mental health care in California isn't just scarce — it's unevenly scarce. A provider might be listed in San Francisco at $200 a session and in a rural county at $168, but the rural patient has a fraction of the income. That gap between what care costs and what people can actually pay is what this analysis is about.

Using the [Two Dimensions of Access dataset](https://osf.io/6ezrx) — which draws on Psychology Today provider listings across U.S. counties — this analysis examines two dimensions of access across California's 58 counties:

- **Availability (AVI):** How many providers exist per 100,000 residents
- **Affordability (AFI):** What 10 therapy sessions costs as a share of disposable income

The findings challenge a common assumption: that rural areas struggle only with availability. In California, rural counties have the highest median fees *and* the lowest incomes — making affordability the sharper edge of the access problem.

---

## Data & Methods

```{r load-data}
# Update this path to your local copy of the OSF dataset
# Download from: https://osf.io/6ezrx
DATA_PATH <- here::here("data", "Two Dimension of access.csv")

raw <- read_csv(DATA_PATH, show_col_types = FALSE)
cat("Full dataset:", nrow(raw), "rows ×", ncol(raw), "columns\n")
```

```{r prepare-california}
ca_all <- raw |>
  filter(State == "CA") |>
  mutate(
    urban_rural = case_when(
      UIC_2024 %in% 1:4 ~ "Urban",
      UIC_2024 %in% 5:6 ~ "Micropolitan",
      UIC_2024 %in% 7:9 ~ "Rural"
    ),
    urban_rural = factor(urban_rural, levels = c("Urban", "Micropolitan", "Rural")),
    fips = str_pad(as.character(county_fips_str), 5, pad = "0"),
    fee_burden = if_else(
      !is.na(median_fee) & !is.na(disposable_income) & disposable_income > 0,
      (median_fee * 10) / disposable_income * 100,
      NA_real_
    ),
    afi_wins = AFI_county_wins_05,
    avi      = AVI_county
  )

ca <- ca_all |> filter(!is.na(median_fee))

# Reference values
national_median_fee <- median(raw$median_fee, na.rm = TRUE)
national_no_prov    <- mean(is.na(raw$median_fee)) * 100

cat("California counties (total):", nrow(ca_all), "\n")
cat("Provider counties:", nrow(ca), "\n")
cat("Excluded (no provider):", nrow(ca_all) - nrow(ca), "\n")
cat("National median fee: $", national_median_fee, "\n")
```

**Sample:** 57 of California's 58 counties have at least one listed provider. The one excluded county is identified in the reproducibility check at the end.

**Fee burden** is calculated as: `(median session fee × 10) / disposable income × 100`. This represents the share of annual disposable income a resident would spend on ten therapy sessions — a commonly used threshold for a basic course of treatment.

---

## Insight 1 — The Affordability Paradox

Rural counties charge the highest median session fee ($167.50) despite having the lowest disposable incomes. The result is that the 10-session burden in non-urban counties (4.9%) is meaningfully higher than in urban ones (3.8%) — even though the absolute fee difference is modest.

```{r insight1-summary}
fee_summary <- ca |>
  group_by(urban_rural) |>
  summarise(
    n              = n(),
    median_fee     = median(median_fee, na.rm = TRUE),
    median_income  = median(disposable_income, na.rm = TRUE),
    mean_burden    = mean(fee_burden, na.rm = TRUE),
    median_avi     = median(avi, na.rm = TRUE),
    mean_medicaid  = mean(pct_medicaid, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

```{r table1, echo=FALSE}
fee_summary |>
  select(urban_rural, median_fee, median_income, mean_burden,
         mean_medicaid, median_avi) |>
  rename(
    "Type"                = urban_rural,
    "Median Fee ($)"      = median_fee,
    "Median Income ($)"   = median_income,
    "Avg Burden (%)"      = mean_burden,
    "Medicaid Accept (%)" = mean_medicaid,
    "Median AVI"          = median_avi
  ) |>
  gt() |>
  tab_header(
    title    = "Table 1. Financial Access Metrics by Urban-Rural Type",
    subtitle = "California provider counties only (n = 57)"
  ) |>
  fmt_currency(columns = 2:3, decimals = 0) |>
  fmt_number(columns = 4:5, decimals = 1, suffix = "%") |>
  fmt_number(columns = 6, decimals = 1) |>
  tab_source_note("Source: Two Dimensions of Access dataset (OSF, 2024–2025)")
```

The paradox here isn't subtle. A rural patient paying $167.50 per session on $34,000 of disposable income is in a structurally worse position than an urban patient paying $175 on $52,000. Fee comparisons alone obscure this.

---

## Insight 2 — The Medicaid Metro Gap

Urban areas dominate provider counts, but their Medicaid acceptance rate is just **1.3%** — functionally zero. Micropolitan counties, despite having far fewer providers in total, accept Medicaid at a rate ten times higher (13.0%).

```{r insight2-chart, fig.width=9, fig.height=5.5}
ins_summary <- ca |>
  group_by(urban_rural) |>
  summarise(
    Insurance = mean(pct_insurance_acceptance, na.rm = TRUE),
    Medicaid  = mean(pct_medicaid, na.rm = TRUE),
    Sliding   = mean(pct_sliding_scale, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))

ins_long <- ins_summary |>
  pivot_longer(-urban_rural, names_to = "Metric", values_to = "pct") |>
  mutate(Metric = factor(Metric, levels = c("Insurance", "Medicaid", "Sliding")))

p_ins <- ggplot(ins_long, aes(x = urban_rural, y = pct, fill = Metric)) +
  geom_col(position = position_dodge(0.75), width = 0.65,
           colour = "white", linewidth = 0.3) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    position = position_dodge(0.75), vjust = -0.4,
    size = 3.5, fontface = "bold"
  ) +
  scale_fill_manual(
    values = c(Insurance = "#2563EB", Medicaid = "#DC2626", Sliding = "#7C3AED"),
    name = NULL
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, "%"),
    limits = c(0, 80), expand = c(0, 0)
  ) +
  labs(
    title    = "Figure 1. Insurance, Medicaid & Sliding-Scale Acceptance",
    subtitle = "California provider counties. Urban Medicaid acceptance is 1.3%.",
    x = NULL, y = "Acceptance Rate (%)",
    caption  = "Source: Two Dimensions of Access dataset, 2024–2025"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(colour = "grey40"),
    legend.position    = "bottom",
    panel.grid.major.x = element_blank()
  )

ggplotly(p_ins, tooltip = c("x", "fill", "y")) |>
  layout(legend = list(orientation = "h", y = -0.2))
```

This matters for policy: Medicaid expansion and network adequacy rules are most often applied to urban markets, where uptake is already the lowest.

---

## Insight 3 — Fee Burden vs. Income

The scatter plot below shows every California provider county. The 10-session burden ranges from under 2% in high-income urban counties to nearly **10% in Sierra County** — where a low income collides with above-average fees.

```{r insight3-scatter, fig.width=10, fig.height=6.5}
top5 <- ca |>
  filter(!is.na(fee_burden)) |>
  slice_max(fee_burden, n = 5)

plot_ly(
  data = ca |> filter(!is.na(fee_burden)),
  x = ~disposable_income,
  y = ~fee_burden,
  color = ~urban_rural,
  colors = pal,
  type  = "scatter",
  mode  = "markers",
  marker = list(size = 10, opacity = 0.7),
  text = ~paste0(
    "<b>", County_Name, "</b><br>",
    "Income: $", format(round(disposable_income), big.mark = ","), "<br>",
    "Burden: ", round(fee_burden, 1), "%<br>",
    "Fee: $", round(median_fee, 0)
  ),
  hoverinfo = "text"
) |>
  add_annotations(
    data = top5,
    x    = ~disposable_income,
    y    = ~fee_burden,
    text = ~County_Name,
    showarrow  = TRUE,
    arrowhead  = 2,
    arrowsize  = 0.8,
    arrowcolor = "grey40",
    ax   = 30,
    ay   = -35,
    font = list(size = 11, color = "grey20")
  ) |>
  layout(
    title = list(
      text = "Figure 2. 10-Session Therapy Burden vs. Disposable Income",
      font = list(size = 14)
    ),
    xaxis = list(title = "Disposable Income ($)", tickformat = "$,.0f"),
    yaxis = list(title = "10-Session Burden (% of Income)", ticksuffix = "%"),
    legend = list(orientation = "h", y = -0.2),
    hoverlabel = list(bgcolor = "white")
  )
```

Counties below $25,000 in disposable income consistently spend 6–10% of that income on a basic course of therapy. Hover over any point to see county-level detail.

---

## Insight 4 — Access Deserts (Maps)

The choropleth maps below show where availability and affordability break down geographically. Gray counties have no listed provider.

```{r maps, fig.width=14, fig.height=8, message=FALSE}
ca_sf <- counties(state = "CA", cb = TRUE, resolution = "5m", year = 2022) |>
  st_transform(4326)

ca_map <- ca_sf |>
  left_join(
    ca_all |> select(fips, urban_rural, avi, afi_wins, fee_burden,
                     median_fee, disposable_income, pct_medicaid),
    by = c("GEOID" = "fips")
  ) |>
  mutate(
    has_provider = !is.na(median_fee),
    avi_plot     = if_else(has_provider, avi, NA_real_),
    afi_plot     = if_else(has_provider, afi_wins, NA_real_)
  )

p_avi <- ggplot(ca_map) +
  geom_sf(aes(fill = avi_plot), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(
    palette = "Blues", direction = 1, na.value = "#E5E7EB",
    name = "AVI\n(per 100K)",
    labels = number_format(accuracy = 1)
  ) +
  labs(
    title    = "Availability Index (AVI)",
    subtitle = "Darker = more providers per capita. Gray = no listed provider."
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(colour = "grey40", hjust = 0.5, size = 9)
  )

p_afi <- ggplot(ca_map) +
  geom_sf(aes(fill = afi_plot), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(
    palette = "Oranges", direction = 1, na.value = "#E5E7EB",
    name = "AFI\n(% income)",
    labels = \(x) paste0(round(x, 1), "%")
  ) +
  labs(
    title    = "Affordability Index (AFI)",
    subtitle = "Darker = higher burden. Winsorised at 95th percentile."
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(colour = "grey40", hjust = 0.5, size = 9)
  )

(p_avi + p_afi) +
  plot_annotation(
    title   = "Figure 3. California Mental Health Access by County",
    caption = "Source: Two Dimensions of Access dataset, 2024–2025",
    theme   = theme(
      plot.title   = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.caption = element_text(colour = "grey50")
    )
  )
```

Access deserts cluster in the northern interior and parts of the Central Valley — precisely the counties with the lowest incomes and fewest alternative resources.

---

## Insight 5 — Provider Availability Distribution

```{r avi-boxplot, fig.width=8, fig.height=5.5}
p_avi_box <- ggplot(ca, aes(x = urban_rural, y = avi, fill = urban_rural)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 2,
              aes(colour = urban_rural)) +
  scale_fill_manual(values = pal) +
  scale_colour_manual(values = pal) +
  labs(
    title    = "Figure 4. Provider Availability by Urban-Rural Type",
    subtitle = "Micropolitan counties have the lowest median AVI.",
    x = NULL, y = "AVI (providers per 100,000)",
    caption  = "Source: Two Dimensions of Access dataset, 2024–2025"
  ) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey40"),
    legend.position = "none"
  )

ggplotly(p_avi_box, tooltip = c("y", "x")) |>
  layout(showlegend = FALSE)
```

Micropolitan counties — mid-sized towns not classified as urban — have the lowest median AVI. This suggests a trough effect: rural counties sometimes benefit from small-town practitioners, while micropolitan areas fall between the coverage of urban networks and the self-sufficiency of rural generalists.

---

## Reproducibility Check

```{r repro-check}
cat("CA total counties:            ", nrow(ca_all), "\n")
cat("CA provider counties:         ", nrow(ca), "\n")
cat("CA excluded (no provider):    ", nrow(ca_all) - nrow(ca), "\n")

excluded <- ca_all |> filter(is.na(median_fee))
cat("Excluded county:              ", excluded$County_Name,
    "(pop", format(excluded$population, big.mark = ","),
    ", UIC", excluded$UIC_2024, ")\n")

cat("CA median session fee:        $", median(ca$median_fee, na.rm = TRUE), "\n")
cat("National median session fee:  $", national_median_fee, "\n")
cat("CA fee premium:               ",
    round((median(ca$median_fee, na.rm = TRUE) / national_median_fee - 1) * 100, 0), "%\n")
cat("Urban Medicaid acceptance:    ",
    round(mean(ca$pct_medicaid[ca$urban_rural == "Urban"], na.rm = TRUE), 1), "%\n")
cat("Micro Medicaid acceptance:    ",
    round(mean(ca$pct_medicaid[ca$urban_rural == "Micropolitan"], na.rm = TRUE), 1), "%\n")
cat("Rural Medicaid acceptance:    ",
    round(mean(ca$pct_medicaid[ca$urban_rural == "Rural"], na.rm = TRUE), 1), "%\n")
cat("Max burden county:            ",
    ca |> filter(!is.na(fee_burden)) |>
      slice_max(fee_burden, n = 1) |> pull(County_Name),
    "(",
    ca |> filter(!is.na(fee_burden)) |>
      slice_max(fee_burden, n = 1) |> pull(fee_burden) |> round(1), "%)\n")
```

---

## Data Source

**Two Dimensions of Access** (OSF, v1.0, January 2026)  
Provider data sourced from the Psychology Today therapist directory.  
Dataset available at: [https://osf.io/6ezrx](https://osf.io/6ezrx)

To reproduce this analysis:

1. Download the dataset from OSF
2. Place the CSV in a `data/` folder in this project directory
3. Run `quarto render index.qmd`

---

```{r session-info, code-fold=true}
sessionInfo()
```
